---
title: "BayesSpace"
author: "Edward Zhao, Xing Ren, Matt Stone, and Raphael Gottardo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BayesSpace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(BayesSpace)
library(ggplot2)
```

## Preprocessing

We'll obtain one of the melanoma samples from spatialLIBD for this example.
```{r}
sce <- spatialLIBD::fetch_data(type="sce", destdir="~/Downloads/spatialLIBD/")
sce <- sce[, sce$sample_name == "151673"]
```

The spatialLIBD samples already include PCs, but in general we need to
preprocess our samples by running PCA.
```{r}
# Dim reduction and clustering
set.seed(102)
dec <- scran::modelGeneVarByPoisson(sce)
top <- scran::getTopHVGs(dec, prop=0.1)
sce <- scran::denoisePCA(sce, technical=dec, subset.row=top)
```

## Clustering
The `spatialCluster()` function will cluster the spots, and add the predicted
cluster labels to `colData(sce)`.
```{r}
sce <- spatialCluster(sce, q=4, init.method="kmeans", model="t", nrep=1000, save.chain=TRUE)
head(sce$spatial.cluster)
```

We can plot the cluster assignments over the spatial locations of the spots.
```{r}
ggplot(as.data.frame(colData(sce)), aes(x = col, y = row, fill = factor(spatial.cluster))) +
  geom_point(size = 7, pch = 22) +
  labs(x = NULL, y = NULL, fill = "Cluster") +
  scale_fill_manual(values = c("purple", "red", "blue", "yellow")) +
  theme_void() + 
  coord_fixed()
```
## Enhanced resolution

The `spatialCluster()` function will enhance the resolution of the principal
components, and add these PCs as well as predicted cluster labels at subspot
resolution to a new SingleCellExperiment.
```{r}
enhanced <- spatialEnhance(sce, q=4, init.method="kmeans", model="t", nrep=1000, save.chain=TRUE)
head(enhanced$spatial.cluster)
```

We can plot the enhanced cluster assignments as above.
```{r}
ggplot(as.data.frame(colData(enhanced)), aes(x = col, y = row, fill = factor(spatial.cluster))) +
  geom_point(size = 3, pch = 22) +
  labs(x = NULL, y = NULL, fill = "Cluster") +
  scale_fill_manual(values = c("purple", "red", "blue", "yellow")) +
  theme_void() + 
  coord_fixed()
```

Note that the enhanced PCs are stored as a `reducedDim` and the enhanced
SingleCellExperiment has no assays. To impute expression (or other features) at
subspot resolution, we need to run `enhanceFeatures()` afterwards.

By default, log-normalized expression (`logcounts(sce)`) is imputed, but other
assays or arbitrary features can be specified.
```{r}
enhanced <- enhanceFeatures(enhanced, sce) 
```

TODO: add example of running SPOTlight and imputing proportion vectors

## Accessing Markov chains

If `save.chain` is set to `TRUE` in either `spatialCluster()` or
`spatialEnhance()`, the chain associated with the respective MCMC run is
preserved to disk as an HDF5 file. The path to this file is stored in the
SingleCellExperiment's metadata at `metadata(sce)$h5.chain`, and can be read
directly using `mcmcChain()`.

The chain is provided as a `coda::mcmc` object, which can be analyzed with
[TidyBayes](https://mjskay.github.io/tidybayes/) or as a matrix. The object has
one row per iteration, with the values of the parameters concatenated across the
row. Columns are named with the parameter name and index (if any).
```{r}
chain <- mcmcChain(sce)
chain[1:5, 1:5]
```

To remove the HDF5 file from disk and remove its path from the metadata, use
`removeChain()`.